Wastewater Surveillance for SARS-CoV-2 Data Analysis Project

### 2023-05-05 code/modeling_code/4_AssayPos_7DMA_model.Rmd
manuscript update

```{r}
aug_full %>% ggplot(aes(x=date)) +
 stat_smooth(aes(y=case_7dma), method = "loess", span = 0.1, color = "black", fill = "black") +
  stat_smooth(aes(y=.pred), method = "loess", span = 0.1, color = "blue", fill = "blue") +
  theme_light()
  ggsave(here("figures/modeling/assaypos_7dma."))
```

>>>>

```{r}
aug_full %>% ggplot(aes(x=date)) +
 stat_smooth(aes(y=case_7dma), method = "loess", span = 0.1, color = "black", fill = "black") +
  stat_smooth(aes(y=.pred), method = "loess", span = 0.1, color = "blue", fill = "blue") +
  theme_light() +
  xlab(label = "Date") +
  ylab(label = "Clinical Case 7-Day Moving Average")

  ggsave(here("figures/modeling/assaypos_7dma.png"))
```

### 2023-05-05 products/manuscript/Manuscript.qmd
manuscript update

At the time of writing it has been three years since the COVID-19 Pandemic was declared, followed by a lockdown across the globe. To date, the World Health Organization reports over 760 million confirmed cases and 6.9 million deaths globally [@WHOCoronavirusCOVID19]. The causative agent, severe acute respiratory syndrome coronavirus-2 (SARS-CoV-2), is a positive-sense single-stranded RNA zoonotic virus that infects cells via its spike proteins or "crowns" [@vkovskiCoronavirusBiologyReplication2021]. When infected by this airborne pathogen, people report experiencing a wide range of mild to severe symptoms, including shortness of breath, cough, head and body aches, and loss of taste or smell. Older adults and others with underlying issues are at higher risk of experiencing more severe outcomes, including death [@COVID19YourHealth2020]. However, long-term effects of COVID-19 on the body are not well understood yet and it may be years before enough research has been done to well-characterize these impacts. The effects of this pandemic have been far-reaching, with influence practically in every aspect of society, indicating a need to keep a close eye on this infectious disease. However, as the pandemic progresses, clinical reporting continues to dwindle, leaving researchers needing a better means of collecting data on the spread of the virus.

@@

Our lab

@@

24-hour composite raw wastewater influent samples were collected twice-weekly from three treatment facilities in Clarke County and stored at 4°C until ready for extraction. 280 𝜇L of wastewater in replicates (n=6) was aliquoted and direct RNA extraction (Zymo Research, R2042) was performed on the same day as sample collection.Each extraction replicate was reverse transcribed and amplified using RT-qPCR and a SARS-CoV-2 specific primer-probe panel (IDT, 10006713), done in technical replicates (n=3). As part of the qPCR panel, two gene targets on the N-gene of the SARS-CoV-2 genome were used for amplification. The N1 and N2 gene targets were chosen due to their conservation across variants and act as universal detectors of the virus \[\@\]. Cycle threshold (Ct) values were then calculated for each PCR reaction, which were then used to estimate SARS-CoV-2 viral load (See Section 3.2 for more detail).

@@

INSERT EQUATION 1

WWTP data was then transformed by converting the flow rate from millions of gallons per day to liters per day. qPCR data was then combined with WWTP data and total viral load per sample was calculated by multiplying average copies per liter by the flow rate. Average viral load for each sample was then determined by taking the mean of the viral load for both N1 and N2 gene targets. County-level daily viral load was calculated by taking the sum of the average viral load from each of the three WWTP samples.

@@

```{r,  echo=FALSE, message=FALSE, fig.cap="Figure 2. Clinical case 7-day moving average time-series"}

@@

@@

The strength of this study lies in the ability of multiple wastewater metrics to predict clinical case data and that an alternative metric to viral load was identified. However, if improvements were to be made, more model tuning would be appropriate, seeing as certain model performance metrics did not far-excede the null model's performance. It is also worth noting that this could be in part due to the maxing out of assay percent positivity, making it difficult for the model to accurately estimate case abundance once the assay has become completely saturated.

@@

Overall, the content of this work highlights the efficacy of these wastewater surveillance methods in capturing accurate trends in clinical cases within this county. Also, qPCR assay positivity was determined to be a slightly better predictor of clinical data than wastewater viral load. Subsequently, a linear regression model was trained on qPCR assay positivity to predict clinical seven-day moving average and was applied to a time-series of low case reporting. Due to upper-bound limitations of the model, the highest case levels were under-predicted both in the training and testing data. However, it is worth noting that during the Fall of 2022 when case reporting was low but assay positivity was high, the model estimated that case levels were much higher than what was reported.

>>>>

At the time of writing it has been three years since the COVID-19 Pandemic was declared, followed by a lockdown across the globe. To date, the World Health Organization reports over 760 million confirmed cases and 6.9 million deaths globally [@WHOCoronavirusCOVID19]. The causative agent, Severe Acute Respiratory Syndrome Coronavirus-2 (SARS-CoV-2), is a positive-sense single-stranded RNA zoonotic virus that infects cells via its spike proteins or "crowns" [@vkovskiCoronavirusBiologyReplication2021]. When infected by this airborne pathogen, people report experiencing a wide range of mild to severe symptoms, including shortness of breath, cough, head and body aches, and loss of taste or smell. Older adults and others with underlying issues are at higher risk of experiencing more severe outcomes, including death [@COVID19YourHealth2020]. However, long-term effects of COVID-19 on the body are not well understood yet and it may be years before enough research has been done to well-characterize these impacts. The effects of this pandemic have been far-reaching, with influence practically in every aspect of society, indicating a need to keep a close eye on this infectious disease. However, as the pandemic progresses, clinical reporting continues to dwindle, leaving researchers needing a better means of collecting data on the spread of the virus.

@@

@@

24-hour composite raw wastewater influent samples were collected twice-weekly from three treatment facilities in Clarke County and stored at 4°C until ready for extraction. 280 𝜇L of wastewater in replicates (n=6) was aliquoted and direct RNA extraction (Zymo Research, R2042) was performed on the same day as sample collection.Each extraction replicate was reverse transcribed and amplified using RT-qPCR and a SARS-CoV-2 specific primer-probe panel (IDT, 10006713), done in technical replicates (n=3). As part of the qPCR panel, two gene targets on the N-gene of the SARS-CoV-2 genome were used for amplification. The N1 and N2 gene targets were chosen due to their conservation across variants and act as universal detectors of the virus. Cycle threshold (Ct) values were then calculated for each PCR reaction, which were then used to estimate SARS-CoV-2 viral load (See Section 3.2 for more detail).

@@

$$
C_W=C_R*(V_Q/V_D)*(V_T/V_R)*(V_E/V_W)*(1.0*10^6uL/1L)
$$

*Equation 1. Viral copies per liter of wastewater (C_W) was calculated by first multiplying copies per qPCR reaction (C_R) by volume of reaction (V_Q) divided by volume of converted DNA (V_D) in the reaction, then multiplying by volume of the reverse-transcription reaction (V_R) divided by volume of RNA extract (V_R) in that reaction, then multiplying by the elution volume of the RNA extraction (V_E) divided by the total volume of wastewater input (V_W). This yields copies per microliter of raw wastewater, which is finally converted to copies per liter.*

WWTP data was then transformed by converting the flow rate from millions of gallons per day to liters per day (Equation 2). qPCR data was then combined with WWTP data and total viral load per sample was calculated by multiplying average copies per liter by the flow rate. Average viral load for each sample was then determined by taking the mean of the viral load for both N1 and N2 gene targets. County-level daily viral load was calculated by taking the sum of the average viral load from each of the three WWTP samples.

$$
VL_W=C_W*V_I
$$

*Equation 2. Wastewater viral load (VL_W) was determined by multiplying viral copies per liter of wastewater (C_W) by the total daily volume of influent flow in liters (V_I).*

@@

```{r,  echo=FALSE, message=FALSE, fig.cap="Figure 2. Clinical case 7-day moving average time-series", fig.width=10, fig.height=8}

@@

In Figure 6, 7DMA predicted by the APP model is shown in blue, whereas the reported value is shown in black. The closest predictions appear to take place towards the beginning of the time-series, whereas the gap in predicted vs actual value widens as time progresses. It also appears that in many instances of high case abundance, the predicted values are under-estimated. This is likely because at a certain point, the qPCR assays reach nearly 100% and cannot accurately estimate values above that threshold. However, it is interesting that in the Fall of 2023 when clinical cases appeared to be severely under-reported (Figure 3), cases are estimated to be nearly as high as during the prior peak during the late summer months, whereas case abundance in the reported data is seemingly much lower.

@@

The strength of this study lies in the ability of multiple wastewater metrics to predict clinical case data and that an alternative metric to viral load was identified. However, if improvements were to be made, more model tuning would be appropriate, seeing as certain model performance metrics did not far-exceed the null model's performance. It is also worth noting that this could be partially due to the maxing out of assay percent positivity, making it difficult for the model to accurately estimate case abundance once the assay has become completely saturated. Because of this, this particular model may be more useful at predicting periods of high vs low case abundance rather than trying to accurately estimate case counts. If done again, I would categorize each date in the wastewater time-series as either high, medium, or low viral abundance based on the distribution of the data, and build a categorical prediction model to predict periods of high, medium, or low case abundance.

@@

Overall, the content of this work highlights the efficacy of these wastewater surveillance methods in capturing accurate trends in clinical cases within this county. qPCR assay positivity was also determined to be a slightly better predictor of clinical data than wastewater viral load. Subsequently, a linear regression model was trained on qPCR assay positivity to predict clinical seven-day moving average and was applied to a time-series of low case reporting. Model performance metrics were better when the model was fit to the training data (prior to February 2022), but performance worsened significantly when fit to testing data (after February 2022). Due to upper-bound limitations though, the highest case levels were under-predicted both in the training and testing data. However, it is still worth noting that during the Fall of 2022 when case reporting was low but assay positivity was high, the model estimated that case levels were much higher than what was reported.
